3.6: Regression with Categorical Data

### 2019-11-12 content/class/18-class.Rmd
Add r practice 17 answers

Today we look at data that is categorical.

>>>>

Today we look at how to use data that is categorical (i.e. variables that indicate an observation's membership in a particular group or category). We introduce them into regression models as **dummy variables** that can equal 0 or 1: where 1 indicates membership in a category, and 0 indicates non-membership.

We also look at what happens when categorical variables have more than two values: for regression, we introduce a dummy variable for each possible category - but be sure to leave out one reference category to avoid the dummy variable trap.

### 2019-11-12 content/r/17-r-practice.Rmd
Add r practice 17 answers

@@

- i. Write out *all* of the causal pathways from `X` (treatment of interest) to `Y` (outcome of interest).
- ii. Write at least one testable implication of the model.^[i.e. what variable(s) should be independent of another? or what should be independent of another *once* controlling for other variables?]
- iii. Identify which variable(s) need to be controlled for to estimate the causal effect of `X` on `Y`.
- iv. Write out the regression equation (abstractly) for properly identifying the causal effect, based on part iii.

>>>>

- [<i class="fas fa-external-link-square-alt"></i> `Answer Key (html)`](/r/17-r-practice-answers.html)
- [<i class="fab fa-markdown"></i> `Answer Key (markdown)`](/r/17-r-practice-answers.Rmd)

@@

- i. Write out *all* of the causal pathways from `X` (treatment of interest) to `Y` (outcome of interest).
- ii. Identify which variable(s) need to be controlled for to estimate the causal effect of `X` on `Y`.
- iii. Write out the regression equation (abstractly) for properly identifying the causal effect, based on part ii.

### 2019-11-12 static/r/17-r-practice-answers.Rmd
Add r practice 17 answers

---
title: "3.5: Causal Inference and DAGs - R Practice"
editor_options: 
  chunk_output_type: console
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(pander)
source(file.path(here::here(), "lib", "graphics.R"))
```

@@

- [<i class="fas fa-external-link-square-alt"></i> `Answer Key (html)`](/r/17-r-practice-answers.html)
- [<i class="fab fa-markdown"></i> `Answer Key (markdown)`](/r/17-r-practice-answers.Rmd)

@@

@@

@@

@@

@@

@@

@@

@@

@@

## Question 5
Test each implication given to you by `dagitty.`

@@

@@

@@

@@

@@

@@

@@

## Question 12
Under this new causal model, list *all* of the possible pathways from `cigs` to `bwght`. Which are "front-doors" and which are "back-doors?" Are any blocked by colliders?

## Question 13
Under this new causal model, what is the minimal sufficient set of variables we need to control in order to causally identify the effect of `cigs` on `bwght`?

## Question 14
Estimate the causal effect in this new model by running the appropriate regression. Compare your answers to those in question 8.

## Question 15
Try out drawing this model using the  `ggdag` package in R. See my DAG in question 3 for an example.

>>>>

---
title: "3.5: Causal Inference and DAGs - R Practice (Answers)"
author: "Ryan Safner"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    #theme: 
    toc: true 
    toc_depth: 3
    toc_float: true
    code_folding: show
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

@@

---

i. Pathways:
  1. $X \rightarrow Y$ (causal, front door)
  2. $X \leftarrow Z \rightarrow Y$ (not causal, back door)

ii. $Z$ needs to be controlled for, since it is opening a backdoor path.

iii. $Y=\beta_0+\beta_1X+\beta_2 Z$

![](images/Adagitty.png)

---

@@

---

i. Pathways:
  1. $X \rightarrow Y$ (causal, front door)
  2. $X \rightarrow M \rightarrow Y$ (causal, front door)

ii. Nothing should be controlled for, since $M$ is a mediator, and part of the effect of $X$ on $Y$

iii. $Y=\beta_0+\beta_1X$

![](images/Bdagitty.png)

---

@@

---

i. Pathways:
  1. $X \rightarrow Y$ (causal, front door)
  2. $X \leftarrow A \rightarrow Z \rightarrow Y$ (not causal, back door)
  3. $X \leftarrow A \rightarrow Z \leftarrow B \rightarrow Y$ (not causal, back door)

ii. Backdoor path 3 is closed by the collider at $Z$. Backdoor path 2 remains open, so we need to control for $A$. (If we blocked $Z$ to close path 2, that would open up backdoor path 3!) Only $A$ should be controlled for.

iii. $Y=\beta_0+\beta_1X+\beta_2A$

![](images/Cdagitty.png)

---

@@

---

i. Pathways:
  1. $X \rightarrow Y$ (causal, front door)
  2. $X \rightarrow C \rightarrow Y$ (causal, front door)
  3. $X \leftarrow A \rightarrow Z \rightarrow B \rightarrow Y$ (not causal, back door)

ii. Path 2 is a front door path we want to leave open. Backdoor path 3 is closed by the collider at $Z$. Nothing needs to be controlled for!

iii. $Y=\beta_0+\beta_1X$

![](images/Ddagitty.png)

---

@@

---

i. Pathways:
  1. $X \rightarrow Y$ (causal, front door)
  2. $X \rightarrow Z \rightarrow Y$ (causal, front door)
  3. $X \rightarrow Z \leftarrow A \rightarrow Y$ (not causal, back door)
  4. $X \rightarrow Z \leftarrow B \rightarrow A \rightarrow Y$ (not causal, back door)

ii. Path 2 is a front door path we want to leave open. Backdoor path 3 is closed by the collider at $Z$. Backdoor path 4 is closed by the collider at $Z$. We don't want to control for anything!

iii. $Y=\beta_0+\beta_1X$

![](images/Edagitty.png)

---

@@

---

@@

---

<!--ANSWER BELOW HERE-->

```{r}
bwght %>%
  select(bwght, cigs, motheduc, cigprice, faminc) %>%
  cor(use = "pairwise.complete.obs")
```

---

@@

---

<!--ANSWER BELOW HERE-->

```{r}

```

---

@@

---

![](images/4dagitty.png)

See the middle box on the right on dagitty:

1. $bwght \perp price \, | \, cigs, inc$: birthweight is independent of price, controlling for cigarettes and income

2. $bwght \perp educ \, | \, cigs, inc$: birthweight is independent of education, controlling for cigarettes and income

3. $inc \perp price$: income is independent of price

4. $price \perp educ$: cigarette price is independent of education

---

## Question 5

Test each implication given to you by `dagitty.`

@@

---

### Implication 1:

If we run a regression of `bwght` on `cigprice`, including `cigs` and `faminc` as controls, there should not not be a statistically significant coefficient on `cigprice` (i.e. there is no relationship between `cigprice` and `bwght` holding `cigs` and` faminc` constant):

```{r}
lm(bwght~cigprice+cigs+faminc, data = bwght) %>% summary()
```

The coefficient on `cigprice` is small and not statistically significant. This implication holds up well.

### Implication 2:

If we run a regression of `bwght` on `motheduc`, including `cigs` and `faminc` as controls, there should not not be a statistically significant coefficient on `cigprice` (i.e. there is no relationship between `motheduc` and `bwght` holding `cigs` and` faminc` constant):

```{r}
lm(bwght~motheduc+cigs+faminc, data = bwght) %>% summary()
```

### Implication 3:

The model implies simply that there is no significant correlation between `faminc` and `cigprice`

```{r}
bwght %>%
  select(faminc, cigprice) %>%
  cor()
```

There is a fairly weak correlation. This implication mostly holds up.

### Implication 4:

The model implies simply that there is no significant correlation between `cigprice` and `motheduc`

```{r}
bwght %>%
  select(cigprice, motheduc) %>%
  cor(use="pairwise.complete.obs")
```

This is an even weaker correlation. This implication seems to holds up.

---

@@

---

<!--ANSWER BELOW HERE-->

1. $cigs \rightarrow bwght$ (causal, front door)
2. $cigs \leftarrow faminc \rightarrow bwght$ (non-causal, back door)
3. $cibs \leftarrow motheduc \rightarrow faminc \rightarrow bwght$ (non-causal, back door)

There are no colliders on any path.

---

@@

---

![](images/7dagitty.png)

We simply need to control for `faminc`. This blocks the back door for both path 2 and path 3.

---

@@

---

We need to control only for `faminc`, so we put it into the regression to estimate:

$$bwght_i=\beta_0+\beta_1 \, cigs_i+\beta_2 \,faminc_i$$
<!--ANSWER BELOW HERE-->

```{r}
lm(bwght~cigs+faminc, data = bwght) %>% summary()
```

Controlling for income, each cigarette smoked while pregant will cause the birthweight to decrease by 0.46 ounces.

---

@@

---

<!--ANSWER BELOW HERE-->

![](images/9dagitty.png)

---

@@

---

<!--ANSWER BELOW HERE-->

See the middle box on the right on dagitty:

1. $bwght \perp price \, | \, cigs, inc, educ$: birthweight is independent of price, controlling for cigarettes, income, and education

2. $price \perp educ$: cigarette price is independent of education

![](images/10dagitty.png)

---

@@

---

<!--ANSWER BELOW HERE-->

### Implication 1:

If we run a regression of `bwght` on `cigprice`, including `cigs`, `faminc`, and `motheduc` as controls, there should not not be a statistically significant coefficient on `cigprice` (i.e. there is no relationship between `cigprice` and `bwght` holding `cigs`,` faminc`, and `motheduc` constant):

```{r}
lm(bwght~cigprice+cigs+faminc+motheduc, data = bwght) %>% summary()
```

The coefficient on `cigprice` is small and not statistically significant. This implication holds up well.

### Implication 2:

This is the same as implication 4 from question 5. Again, this holds up reasonably well.

---

## Question 12

Under this new causal model, list *all* of the possible pathways from `cigs` to `bwght`. Which are "front-doors" and which are "back-doors?" Are any blocked by colliders?

---

<!--ANSWER BELOW HERE-->

1. $cigs \rightarrow bwght$ (causal, front door)
2. $cigs \leftarrow faminc \rightarrow bwght$ (non-causal, back door)
3. $cigs \leftarrow cigprice \leftarrow u1 \rightarrow faminc \rightarrow bwght$ (non-causal, back door)
4. $cigs \leftarrow motheduc \rightarrow bwght$ (non-causal, back door)
5. $cigs \leftarrow motheduc \rightarrow faminc \rightarrow bwght$ (non-causal, back door)

There are no colliders on any path.

---

## Question 13

Under this new causal model, what is the minimal sufficient set of variables we need to control in order to causally identify the effect of `cigs` on `bwght`?

---

<!--ANSWER BELOW HERE-->

We need to control for `faminc` and `motheduc`. This blocks the back door for paths 2, 3, 4, and 5.

---

## Question 14

Estimate the causal effect in this new model by running the appropriate regression. Compare your answers to those in question 8.

---

<!--ANSWER BELOW HERE-->

![](images/14dagitty.png)

We need to control for `faminc` and `motheduc`, so we put them into the regression to estimate:

$$bwght_i=\beta_0+\beta_1 \, cigs_i+\beta_2 \,faminc_i+ \beta_3 \, motheduc_i$$
<!--ANSWER BELOW HERE-->

```{r}
lm(bwght~cigs+faminc+motheduc, data = bwght) %>% summary()
```

Controlling for income and education, each cigarette smoked while pregant will cause the birthweight to decrease by 0.46 ounces. It turns out there was no noticeable difference when we included education!

---

## Question 15

Try out drawing this model using the  `ggdag` package in R. See my DAG in question 3 for an example.

---

<!--ANSWER BELOW HERE-->

```{r, echo=T}
library(ggdag)
dagify(bwght~cigs+inc+educ,
       cigs~price+educ+inc,
       inc~educ+u1,
       price~u1,
       exposure = "cigs",
       outcome = "bwght") %>% 
  ggdag(stylized = FALSE, seed=1)+theme_dag_blank()+theme(legend.position = "none")
```
```

### 2019-11-12 static/slides/18-slides.Rmd
Add r practice 17 answers

Code region numerically: $$Region_i=
\begin{cases}1 & \text{if } i \text{ is in }Northeast \\
   			2 & \text{if } i \text{ is in } Midwest\\
    		3 & \text{if } i \text{ is in } South \\
   			4 & \text{if } i \text{ is in } West\\ \end{cases}$$
]

>>>>

Code region numerically: 
$$Region_i=
\begin{cases}1 & \text{if } i \text{ is in }Northeast\\
   			2 & \text{if } i \text{ is in } Midwest\\
    		3 & \text{if } i \text{ is in } South \\
   			4 & \text{if } i \text{ is in } West\\ 
\end{cases}$$
